Structure Learning for $\VAR(p)$ Models

### 2019-05-21 paper/paper.tex
Added some citations and updated intro of paper.

\section{Introduction and Review}
\label{sec:introduction}
Consider a collection of stochastic processes producing observations
at discrete time intervals.  Are the underlying processes dependent?
Can we quantify any of the underlying relationships?  Can the arrow of
time help us to distinguish a directionality or flow of dependence
among our observed series?  In this paper we contribute to the
understanding of the notion of Granger-Causality
\cite{granger1969investigating} as a tool for answering these questions.

Though the notion of causality is a philosophically slippery concept,
it is fundamental to the way we understand the world and to the
progress of science in general.  Indeed, without faith in the
consistency of causal interactions the results of experimental science
could not be generalized or applied in any meaningful way.  In the
case of Granger-Causality, we state that if an event '$A$' provides us
with unique (that is, not available anywhere else) information about a
later event $B$, then $A$ must have a causal impact on $B$.  As
opposed to the notion of Causation promoted by Pearl
\cite{pearl2000art}, this is an entirely model-free notion of cause,
and instead leverages the intuition that a cause must precede it's
effect.

In practice, Granger's notion of causation it not a convincing test
for \textit{true} causation, since our statements about causation are
highly dependent upon the data that we are able to observe.  We prefer
instead to interpret Granger causality as a means of uncovering a flow
of ``information'' or ``energy'' through some underlying graph of
interactions.  Though this graph cannot be observed directly, we will
infer it's presence as a latent structure among our observed time
series data.

Finding the ``best'' graph structure consistent with observed data is
generally an extremely challenging problem, though the comparison of
quality between different structures, and hence the notion of
``best'', needs to be quantified.  In applications where we are
interested merely in minimizing the mean squared error of a linear
one-step-ahead predictor, then we will naturally desire an entirely
dense graph of connections, since each edge can only serve to reduce
estimation error.  However, since the number of edges scales
quadratically in $n$, it is imperative to infer a sparse causality
graph for large systems, both to avoid overfitting observed data, as
well as to aid the interpretability of the results.

In \cite{bach2004learning} the authors apply a local search heuristic
to an AIC penalized approximation of the likelihood where at each
iteration an edge is either added, removed, or reversed.  This is a
common approach to combinatorial optimization due to it's simplicity,
but is liable to get stuck in shallow local minima.

@@

From a Bayesian perspective, spike-and-slab priors can be used to
estimate the probability of an edge's inclusion, after which a graph
can be chosen heuristically by thresholding or searching for the MAP
graph structure.

Our contribution is to provide an alternative heuristic based on
minimum spanning trees, inspired by our result in section
\ref{sec:theory} that the true causality graph can be recovered from
pairwise information alone in the case that the underlying graph is a
directed forest.  We compare this algorithm to a grouped LASSO
\cite{yuan2006model}, local search, and a Bayesian posterior
probability thresholding scheme.

@@

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will denote the collection vectors of
such random elements as $L_2^n(\Omega)$ or simply $L_2^n$.

We will work with a discrete time and wide-sense stationary process
$x(t)$ taking elements in $L_2$.  We suppose that $x(t)$ (with
$t \in \Z$) has zero mean, $\E x(t) = 0$, and the matrix valued
covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ is absolutely
summable and has a spectra $S(\omega)$:

@@

@@

where the closure is naturally in mean-square.  $\H_t^x$ is the space
spanned by the values of $x(t)$ between the ``infinite past'' and
``now'', we will often omit the superscript $x$ which should be clear
from context.  Evidently these spaces are separable, and as closed
subspaces of a Hilbert space they are themselves Hilbert.  We will
denote the spaces generated in analagous ways by particular components
of $x$ as e.g. $\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular
components as e.g. $\H_t^{-j}$.

As a consequence of the Wold decomposition theorem \cite{lindquist},
every such sequence has the ``moving average'' $MA(\infty)$
representation

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) w(t - \tau),
\end{equation}

where $c(t)$ is a perfectly predictable sequence\footnote{$c(t)$ is
  perfectly predictable if any sample $c(t_0)$ is enough to determine
  $c(t)$ for every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, and $w(t)$ is an uncorrelated sequence having a diagonal
covariance and $A(0) = I$.  We will assume that $c(t) = 0$, which in
practice is to say that it has been accurately estimated and removed.

Furthermore, we assume that the spectra is uniformly bounded away from
$0$:
$\exists \delta:\ \delta I \preceq S(\omega) \preceq \frac{1}{\delta}
I\ \forall \omega$ so that the Wold representation can be inverted to
yield the $VAR(\infty)$ form

\begin{equation}
  \label{eqn:ar_representation}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t),
\end{equation}

where $v(t)$ is an uncorrelated sequence with covariance $\Sigma_v$.

The equations (\ref{eqn:wold}), (\ref{eqn:ar_representation}) can be
represented as $x(t) = \A(z)w(t) = \B(z)x(t) + v(t)$ via the action of
the operators (LTI filters)
$\A(z) \defeq \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$ and
$\B(z) \defeq \sum_{\tau = 1}^\infty B(\tau)z^{-\tau}$ where the
operator $z^{-1}$ is the backshift operator acting on $\H$.  Since
$||z^{-1}|| = 1$, the inversion formulae

\begin{equation}
  \label{eqn:lsi_inversion}
  \A(z) = (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k
\end{equation}

are valid.
\subsection{Granger Causality}

@@

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if everything is Gaussian then they are
equivalent \cite{barnett2009granger}.

@@

\begin{definition}[Causality graph, Parents]

Remaining in the context of a collection of $n$ WSS processes
$x(t) = \big(x_1(t), \ldots, x_n(t)\big)$, we define the Granger-causality
graph $\gcg$ to be the directed graph formed on $n$ vertices where an
edge $(j, i) \in \gcg$ if and only if $x_j \gc x_i$.

We refer to the set of \textit{parents} of $x_i$ in $\gcg$ by
$\pa{i}$, that is, $j \in \pa{i} \iff x_j \gc x_i$.
\end{definition}

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with a
\textit{strictly causal} LTI filter

\begin{equation}
  \label{eqn:filter}
  \B_{ij}(z) = \sum_{\tau = 1}^{\infty} B_{ij}(\tau)z^{-\tau}
\end{equation}

from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.  We will represent the action
(convolution) of this filter compactly by taking $z^{-1}$ as the
backshift operator on $\ell_2^n(\Omega, \mathcal{F}, \mathbb{P})$ when
convenient

\begin{equation}
  \B_{ij}(z)x_j(t) \defeq \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau).
\end{equation}

@@

From the $\VAR$ representation of $x(t)$ there is clearly a tight
relationship between each node and it's parent nodes, indeed

@@

% When is this guaranteed w.r.t. the full system?
Furthermore, if the filter $\B_{ii}$ is invertible, we can write

\begin{equation*}
  \begin{aligned}
    (1 - \B_{ii}(z))x_i(t) &= v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\\
    \Rightarrow x_i(t) &= (1 - \B_{ii}(z))^{-1}\big[v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\big]\\
    &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)
  \end{aligned}
\end{equation*}

We continue to develop the relationship between $x_i$ and it's family
tree, where the following proposition is analagous to the definitions
of feedback free processes of \cite{caines1975feedback}.

\begin{proposition}[Ancestor Expansion]
  \label{prop:parent_expanding}
  The component $x_i(t)$ of $x(t)$ can be represented as a sum of
  linear filtering operations acting on the driving noise $v_i$ of
  $x_i$ as well as the parents of $x_i$:

  \begin{equation}
    \label{eqn:parent_expansion}
    x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t),
  \end{equation}

  where $\wtB_{ii}(z) = (1 - \B_{ii}(z))^{-1}$ and
  $\wtB_{ik}(z) = \wtB_{ii}(z) B_{ik}(z)$.

  Moreover, $x_i$ can be expanded in terms of it's ancestors and
  driving noise only:

@@

The matrix $\B(z)$ is analagous to a \textit{weighted adjacency
  matrix} for the graph $\gcg$.  And, in the same way that the
$k^{\text{th}}$ power of an adjacency matrix counts the number of
paths of length $k$ between nodes, $(\B(z)^k)_{ij}$ is a filter
isolating the ``action'' of $j$ on $i$ at a time lag of $k$ steps.
Though we only need the sparsity pattern of $\B(z)^k$ for the proof.

We state the following well known theorem as a lemma, proof follows
easily by induction.

We now prove the proposition.

\begin{proof}
  Equation \eqref{eqn:parent_expansion} was derived prior to the
  statement of the theorem.  We are left to demonstrate equation
  \eqref{eqn:ancestor_expansion}.

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality.

@@

This notion is of interest for a variety of reasons.  From a purely
theoretical or conceptual standpoint, we will seek to understand when
and how pairwise causality can capture the idea of ``flow of
information'' in the underlying graph.  It may also be useful for
reasoning about what the conditions under which \textit{unobserved}
components of $x(t)$ may or may not interfere with inference in the
actually observed components.  Finally, motivated from a practical
standpoint to analyze causation in large systems, we will seek to
construct practical estimation procedures based purely on pairwise
causality tests which can result in reduced computational complexity.

\begin{proposition}
  \label{prop:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  Then, if

@@

  then $H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall \tau \in \Z_+\ \E[x_i(t)x_j(t - \tau)] = 0$.  Moreover,
  this implies that $j \npwgc i$.
\end{proposition}

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the
  \textit{present} of $x_i(t)$ is uncorrelated with $x_j(t)$,
  but we need not have $\H_t^{(i)} \perp \H_t^{(j)}$ since
  $i \in \anc{j}$ is not excluded.

  We also point out the contrapositive that if $j \pwgc i$ then we
  must have either $j \in \anc{i}$ or $\anc{i} \cap \anc{j} \ne 0$.
\end{remark}

@@

  which follows from Proposition \ref{prop:ancestor_uncorrelated}.
  Finally, for $k \in C_1(j)$ the case $k = i$ is immediate (since the
  error in estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given
  $\H_{t - 1}^{(i)}$ is $0$), so suppose $k \ne i$.  We know from
  above that $k$ inherits the key properties referred to as $a, b$ and
  therefore we can recursively expand $k$ in the same way as in
  equation (\ref{eqn:xj_partition_expansion}).  Continuing this
  recursion for each $k \in C_1(j)$ (where $k \ne i$) must eventually
  terminate since $i \in \anc{k}$.
\end{proof}

@@

In this section and the next we will seek to understand when converse
statements of Proposition \ref{prop:ancestor_properties} \textit{do}
hold.  One possibility is to restrict the coefficients of the system
matrix, e.g. by requiring that $B_{ij}(\tau) \ge 0$.  Instead,
we think it is more meaningful to focus on the defining feature of
time series networks, that is, the topology of $\gcg$.

@@

\begin{definition}[Depth]
  For our present purposes we will define the \textit{depth} $d(j)$ of
  a node $j$ in $\gcg$ to be the length of the \textit{longest} path
  from a node in $P_0$ to $j$, where $d(j) = 0$ if $j \in P_0$.  It is
  apparent (\hl{right?}) that such a path will always exist in a
  strongly causal graph.  For example, in Figure
  \ref{fig:example_fig3} we have $d(3) = 1$ and $d(4) = 2$.
\end{definition}

@@

@@

  \BlankLine
  \caption{Filter Estimator}
  \label{alg:Bz_estimate}
  \Input{Wide sense stationary process $x(t)$\\
    Maximum iterations $M$\\
    Sparsity parameters $\{(\delta_m\, b_m, R_m)\}_{m = 1}^M$}
  \Output{Linear one-step-ahead estimator $\widehat{\B}(z)$}
  \Initialize{$\epsilon^{(0)}(t) = x(t)$ \texttt{\# initial error process}\\
    $m = 1$ \texttt{\# iteration counter}}
  \BlankLine
  \For{$m = 1, \ldots, M$}{
    $F \leftarrow \mathsf{pwgc}(\epsilon^{(m - 1)}(t))$ \texttt{\# estimate pairwise Granger-causality statistics}\\
    $\widehat{\gcg}_m \leftarrow \mathsf{pw\_scg}(F, \delta_m, b_m, R_m)$ \texttt{\# graph estimate (Algorithm \ref{alg:pwgr_heuristic})}\\
    $\widehat{\B}^{(m)}(z) \leftarrow \mathsf{estimate\_B}(\epsilon^{(m - 1)}(t), \widehat{\gcg}_m)$ \texttt{\# Estimate filter with topology} $\widehat{\gcg}_m$\\
    $\epsilon^{(m)} \leftarrow \epsilon^{(m - 1)}(t) - \widehat{\B}^{(m)}(z)\epsilon^{(m - 1)}(t)$ \texttt{\# updated error process}\\
  }
  \BlankLine
  $\widehat{\B}(z) \leftarrow \widehat{\B}^{(M)}(z) \prod_{k = 1}^{M - 1}\big(I - \widehat{\B}^{(M - k)}(z)\big)$ \texttt{\# compute total filter}\\
  \Return{$\widehat{\B}(z)$}
\end{algorithm}

\section{Application}
\section{Conclusion}

>>>>

\section{Introduction and Review}
\label{sec:introduction}
Consider a collection of stochastic processes producing observations
at discrete time intervals.  Are the underlying processes dependent?
Can we quantify any of the underlying relationships?  Can the arrow of
time help us to distinguish a directionality or flow of dependence
among our observed series?  In this paper we contribute to the
understanding of the notion of Granger-Causality
\cite{granger1969investigating} \cite{Granger1980329} as a tool for
answering these questions.

Heuristically, we state that if an event '$A$' occuring strictly prior
in time to an even '$B$' and $A$ provides us with exclusive (that is,
not available anywhere else) information about $B$, then $A$ must have
had a causal impact on $B$.  This is an entirely model-free notion of
cause, and instead leverages the intuition that a cause must precede
it's effect.

In practice, Granger's notion of causation it not a convincing test
for \textit{true} causation, since our statements about causation are
highly dependent upon the data that we are able to observe.  Rather,
the interpretation of Granger causality is as a means of uncovering a
flow of ``information'' or ``energy'' through some underlying graph of
interactions.  Though this graph cannot be observed directly, we will
infer it's presence as a latent structure among our observed time
series data.  This intuition is levereged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007}; and power system design
\cite{Misyrlis2016450}, \cite{yuan2014root}.

Finding the ``best'' graph structure consistent with observed data is
generally an extremely challenging problem (i.e. this is a best subset
selection problem, see \cite{bss_mio}, \cite{hastie_bss_comp}), though
the comparison of quality between different structures, and hence the
notion of ``best'', needs to be quantified.  In applications where we
are interested merely in minimizing the mean squared error of a linear
one-step-ahead predictor, then we will naturally desire an entirely
dense graph of connections, since each edge can only serve to reduce
estimation error.  However, since the number of edges scales
quadratically in $n$, it is imperative to infer a sparse causality
graph for large systems, both to avoid overfitting observed data, as
well as to aid the interpretability of the results.

In \cite{bach2004learning} the authors apply a local search heuristic
to the Whittle likelihood with an AIC penalization.  The local search
heuristic where at each iteration an edge is either added, removed, or
reversed is a common approach to combinatorial optimization due to
it's simplicity, but is liable to get stuck in shallow local minima.

@@

The contributions of this paper are as follows: firstly, in section
\ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as a characterization of the sparsity pattern of matrices arising
in the Wold decomposition.  We establish sufficient conditions
(sections \ref{sec:strongly_causal_graphs},
\ref{sec:persistent_systems}) under which a fully conditional
Granger-causality graph can be recovered from pairwise tests alone
(section \ref{sec:pairwise_algorithm}).  Secondly, we propose in
section \ref{sec:structure_learning} alternative graph search
heuristics inspired by our theoretical results.  Our heuristics are
compared against a standard grouped LASSO algorithm
\cite{yuan2006model} as well as local search hill climbing (\hl{state
  here the results}).  In section \ref{sec:application} we develop an
example application (\hl{more detail...}).  And provide concluding
remarks on further open problems in our conclusion (section
\ref{sec:conclusion}).

@@

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) vector valued process $x(t)$ (with
$t \in \Z$) with elements taking values in $L_2$.  We suppose that
$x(t)$ has zero mean, $\E x(t) = 0$, and has absolutely summable
matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ with an absolutely
continuous spectral density $S(\omega)$:

@@

% and that the spectra is bounded uniformly away from 0:

% \begin{equation*}
%   \int_{-\pi}^\pi \ln\det S(\omega) \d\omega > -\infty.
% \end{equation*}

@@

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analagous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular components as
e.g. $\H_t^{-j}$.

As a consequence of the Wold decomposition theorem \cite{lindquist},
every WSS sequence has the moving average $MA(\infty)$
representation

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) v(t - \tau),
\end{equation}

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, and $w(t)$ is an uncorrelated sequence and $A(0) = I$.  We
will assume that $c(t) = 0$, which in practice is to say that it has
been accurately estimated and removed.  We additionally require that
this representation can be inverted to yield the $VAR(\infty)$ form

\begin{equation}
  \label{eqn:ar_representation}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t).
\end{equation}

The equations (\ref{eqn:wold}), (\ref{eqn:ar_representation}) can be
represented as $x(t) = \A(z)v(t) = \B(z)x(t) + v(t)$ via the action
(convolution) of the operators (LTI filters)
$\A(z) \defeq \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$ and
$\B(z) \defeq \sum_{\tau = 1}^\infty B(\tau)z^{-\tau}$ where the
operator $z^{-1}$ is the backshift operator acting on
$\ell_2^n(\Omega, \mathcal{F}, \mathbb{P})$, that is:

\begin{equation}
  \label{eqn:filter_action}
  \B_{ij}(z)x_j(t) \defeq \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau).
\end{equation}

Finally, since $||z^{-1}|| = 1$ we have the inversion formula

\begin{equation}
  \label{eqn:lsi_inversion}
  \A(z) = (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
\end{equation}

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal matrix,
which is referred to as a lack of instantaneous feedback in $x(t)$.
We formally state our setup to conclude this section:

\begin{definition}[Basic Setup]
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal covariance matrix.  The $MA(\infty)$
  representation of equation \eqref{eqn:wold} has $c(t) = 0$ and
  $A(0) = I$.
\end{definition}

\subsection{Granger Causality}

@@

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if the distribution of $v(t)$ is known to be
Gaussian then they are equivalent \cite{barnett2009granger}.

@@

\begin{definition}[Causality graph, Parents]
  We define the Granger-causality graph $\gcg = ([n], \gcge)$ to be the directed
  graph formed on $n$ vertices where an edge $(j, i) \in \gcge$ if and
  only if $x_j$ Granger-causes $x_i$ (with respect to $x$).  That is,
  $$(j, i) \in \gcge \iff j \in \pa{i} \iff x_j \gc x_i.$$
\end{definition}

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with the
\textit{strictly causal} LTI filter $\B_{ij}(z)$ (see eqn
\eqref{eqn:filter_action}).  Thence, the matrix $\B(z)$ is analagous
to a \textit{weighted adjacency matrix}\footnote{We are using the
  convention that $\B_{ij}(z)$ is a filter with input $x_j$ and output
  $x_i$ so as to write the action of the system as $\B(z)x(t)$ with
  $x(t)$ as a column vector.  This competes with the usual convention
  for adjacency matrices where $A_{ij} = 1$ if there is an edge
  $(i, j)$.  In our case, the sparsity pattern of $\B_{ij}$ is the
  \textit{transposed} adjacency matrix.} for the graph $\gcg$.  And,
in the same way that the $k^{\text{th}}$ power of an adjacency matrix
counts the number of paths of length $k$ between nodes,
$(\B(z)^k)_{ij}$ is a filter isolating the ``action'' of $j$ on $i$ at
a time lag of $k$ steps, this is exemplified in the inversion formula
\ref{eqn:lsi_inversion}.

An elementary theorem connecting the adjacency matrix with paths will
allow us to deduce the sparsity pattern of $\A(z)$.  We omit proof,
though it follows easily by induction:

@@

From the $\VAR$ representation of $x(t)$ there is clearly a tight
relationship between each node and it's parent nodes, the following
proposition is analagous to the definition of feedback free processes
of \cite{caines1975feedback}:

% % When is this guaranteed w.r.t. the full system?
% Furthermore, if the filter $\B_{ii}$ is invertible, we can write

% \begin{equation*}
%   \begin{aligned}
%     (1 - \B_{ii}(z))x_i(t) &= v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\\
%     \Rightarrow x_i(t) &= (1 - \B_{ii}(z))^{-1}\big[v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\big]\\
%     &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)
%   \end{aligned}
% \end{equation*}

% We continue to develop the relationship between $x_i$ and it's family
% tree, where the following proposition is analagous to the definitions
% of feedback free processes of \cite{caines1975feedback}.

\begin{proposition}[Ancestor Expansion]
  \label{prop:parent_expanding}
  The component $x_i(t)$ of $x(t)$ can be represented in terms of it's
  parents in $\gcg$:

@@

  Moreover, $x_i$ can be expanded in terms of it's ancestor's $v(t)$
  components only:

@@

\begin{proof}
  Equation \eqref{eqn:parent_expansion} is immediate from the
  $VAR(\infty)$ representation of \eqref{eqn:ar_representation} and
  Theorem \ref{thm:granger_causality_equivalences}, we are left to
  demonstrate \eqref{eqn:ancestor_expansion}.

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In this paper we study
\textit{pairwise} Granger-causality, and seek to understand when
knowledge of pairwise relations is sufficient to deduce the true fully
conditional relations of $\gcg$.

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph.
It may also be useful for reasoning about what the conditions under
which \textit{unobserved} components of $x(t)$ may or may not
interfere with inference in the actually observed components.
Finally, motivated from a practical standpoint to analyze causation in
large systems, we will seek to construct practical estimation
procedures based purely on pairwise causality tests since the
computation of such pairwise relations is somewhat easier.

The following is a fundamental building block:

\begin{lemma}
  \label{lem:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  Then, if

@@

  then $H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall \tau \in \Z_+\ \E[x_i(t)x_j(t - \tau)] = 0$.  Moreover,
  this implies that $j \npwgc i$.
\end{lemma}

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the
  \textit{present} of $x_i(t)$ is uncorrelated with $x_j(t)$,
  but we need not have $\H_t^{(i)} \perp \H_t^{(j)}$ since
  $i \in \anc{j}$ is not excluded.
\end{remark}

@@

  which follows from Lemma \ref{lem:ancestor_uncorrelated}.
  Finally, for $k \in C_1(j)$ the case $k = i$ is immediate (since the
  error in estimating $\Phi(z)\B_{ii}(z)x_i(t)$ given
  $\H_{t - 1}^{(i)}$ is $0$), so suppose $k \ne i$.  We know from
  above that $k$ inherits the key properties referred to as $a, b$ and
  therefore we can recursively expand $k$ in the same way as in
  equation (\ref{eqn:xj_partition_expansion}).  Continuing this
  recursion for each $k \in C_1(j)$ (where $k \ne i$) must eventually
  terminate since $i \in \anc{k}$.
\end{proof}

@@

In this section and the next we will seek to understand when converse
statements of Proposition \ref{prop:ancestor_properties} \textit{do}
hold.  One possibility is to restrict the coefficients of the system
matrix, e.g. by requiring that $B_{ij}(\tau) \ge 0$.  Instead,
we think it more meaningful to focus on the defining feature of
time series networks, that is, the topology of $\gcg$.

@@

\begin{definition}[Depth]
  For our present purposes we will define the \textit{depth} $d(j)$ of
  a node $j$ in $\gcg$ to be the length of the \textit{longest} path
  from a node in $P_0$ to $j$, where $d(j) = 0$ if $j \in P_0$.  It is
  apparent that such a path will always exist.  For example, in Figure
  \ref{fig:example_fig3} we have $d(3) = 1$ and $d(4) = 2$.
\end{definition}

@@

\begin{proposition}
  The error sequence in Algorithm \ref{alg:Bz_estimate} is a zero mean
  WSS process $\E\epsilon^{(m)}(t) = 0$ with monotonically decreasing
  variance: $\E|\epsilon^{(m + 1)}(t)|^2 \le \E|\epsilon^{(m)}(t)|^2$.
\end{proposition}
\begin{proof}
  \hl{TODO -- this is probably not even true}
\end{proof}

@@

  \BlankLine
  \caption{Filter Estimator}
  \label{alg:Bz_estimate}
  \Input{$T$ samples from a WSS process $x: \{x(t)\}_{t = 1}^T$\\
    Maximum iterations $M$\\
    Sparsity parameters $\{(\delta_m\, b_m, R_m)\}_{m = 1}^M$}
  \Output{Linear one-step-ahead estimator $\widehat{\B}(z)$}
  \Initialize{$\epsilon^{(0)}(t) = x(t)$ \texttt{\# initial error process}\\
    $m = 1$ \texttt{\# iteration counter}}
  \BlankLine
  \For{$m = 1, \ldots, M$}{
    $F \leftarrow \mathsf{pwgc}(\epsilon^{(m - 1)}(t))$ \texttt{\# estimate pairwise Granger-causality statistics}\\
    $\widehat{\gcg}_m \leftarrow \mathsf{pw\_scg}(F, \delta_m, b_m, R_m)$ \texttt{\# graph estimate (Algorithm \ref{alg:pwgr_heuristic})}\\
    $\widehat{\B}^{(m)}(z) \leftarrow \mathsf{estimate\_B}(\epsilon^{(m - 1)}(t), \widehat{\gcg}_m)$ \texttt{\# Estimate filter with topology} $\widehat{\gcg}_m$\\
    $\epsilon^{(m)}(t) \leftarrow \epsilon^{(m - 1)}(t) - \widehat{\B}^{(m)}(z)\epsilon^{(m - 1)}(t)$ \texttt{\# updated error process}\\
  }
  \BlankLine
  $\widehat{\B}(z) \leftarrow \widehat{\B}^{(M)}(z) \prod_{k = 1}^{M - 1}\big(I - \widehat{\B}^{(M - k)}(z)\big)$ \texttt{\# compute total filter}\\
  \Return{$\widehat{\B}(z)$}
\end{algorithm}

\section{Application}
\label{sec:application}
\section{Conclusion}
\label{sec:conclusion}
Extensions to transfer entropy?.  Case of time varying statistics?.  Develop sparse estimators for SS models (\cite{barnett2015granger})?.  Apply MIO for the Whittle relaxation?
